---
title: "Winter 2026 B BIO 495"
output: html_notebook
---

Fitness assay analysis
- starting from "long data" --> pair up the correct rows (t0 with t7 for each group / strain pair / predation condition)
- determine change in white colonies. (Blue is just a reference point.)

- This is an R Notebook script -- analysis using the language R, as well as a particular type of data handling called "tidy". "Tidyverse" is a library of functions for reading in, rearranging, and plotting data.


```{r libraries}
library(tidyverse)
library(tidyverse)
library(cowplot)
```

```{r}
# ============================================================
# User settings
# ============================================================
infile = "../datafiles/fitness_assay_all.tsv"
outdir = "../output"
pseudocount = 0
min_colony_count = 20   # minimum total colonies required at BOTH t0 and t6

# Optional filters
keep_blue_strain = "ancB"

# Keep ancW and any white strains starting with "ev"
keep_white_fn = function(x) {
  x == "ancW" | stringr::str_starts(x, "ev")
}

predation_colors = c(
  "yes" = "firebrick2",
  "no"  = "dodgerblue2"
)
```

Read and clean data
```{r}

dat_long =
  readr::read_tsv(infile, show_col_types = FALSE) %>%
  mutate(
    num_blue_colonies  = readr::parse_number(as.character(num_blue_colonies)),
    num_white_colonies = readr::parse_number(as.character(num_white_colonies)),
    timepoint = as.character(timepoint),
    timepoint = if_else(timepoint == "t7", "t6", timepoint),
    predationYN = as.character(predationYN),
    white_strain = gsub(pattern="evPop", replacement="ev", x=white_strain)
  ) %>%
  filter(
    timepoint %in% c("t0", "t6"),
    blue_strain == keep_blue_strain,
    keep_white_fn(white_strain)
  )


dat_long <- dat_long %>%
  mutate(
    strain_num = ifelse(white_strain == "ancW", 0,
                        as.numeric(sub("ev", "", white_strain))),
    white_strain = factor(white_strain,
                          levels = unique(white_strain[order(strain_num)]))
  )
# ============================================================
# Check required columns
# ============================================================
required_cols = c(
  "group_number",
  "white_strain",
  "replicate",
  "predationYN",
  "timepoint",
  "num_blue_colonies",
  "num_white_colonies",
  "blue_strain"
)

missing_cols = setdiff(required_cols, names(dat_long))
if (length(missing_cols) > 0) {
  stop("Missing required column(s): ", paste(missing_cols, collapse = ", "))
}

# ============================================================
# Pair up the t0 and t6 based on the group, white strain ID,
# and replicate (i.e., assay date).
# Note: one t0 row may be reused for more than one t6 row
# ============================================================

pair_keys = c("group_number", "white_strain", "replicate")

dat_t0 =
  dat_long %>%
  filter(timepoint == "t0") %>%
  select(
    all_of(pair_keys),
    blue_strain,
    num_blue_colonies_t0  = num_blue_colonies,
    num_white_colonies_t0 = num_white_colonies
  ) %>%
  distinct()

dat_t6 =
  dat_long %>%
  filter(timepoint == "t6") %>%
  select(
    all_of(pair_keys),
    blue_strain,
    predationYN,
    num_blue_colonies_t6  = num_blue_colonies,
    num_white_colonies_t6 = num_white_colonies
  )

fitness_white_change =
  dat_t6 %>%
  left_join(
    dat_t0,
    by = c("group_number", "white_strain", "replicate", "blue_strain")
  ) %>%
  mutate(
    total_count_t0_raw = num_blue_colonies_t0 + num_white_colonies_t0,
    total_count_t6_raw = num_blue_colonies_t6 + num_white_colonies_t6,

    has_both_timepoints =
      !is.na(num_blue_colonies_t0) &
      !is.na(num_white_colonies_t0) &
      !is.na(num_blue_colonies_t6) &
      !is.na(num_white_colonies_t6),

    passes_min_colonies =
      has_both_timepoints &
      total_count_t0_raw >= min_colony_count &
      total_count_t6_raw >= min_colony_count,

    # Pseudocount-adjusted counts
    blue_t0  = num_blue_colonies_t0  + pseudocount,
    white_t0 = num_white_colonies_t0 + pseudocount,
    blue_t6  = num_blue_colonies_t6  + pseudocount,
    white_t6 = num_white_colonies_t6 + pseudocount,

    total_t0 = blue_t0 + white_t0,
    total_t6 = blue_t6 + white_t6,

    # White frequency at each timepoint
    pW_t0 = white_t0 / total_t0,
    pW_t6 = white_t6 / total_t6,

    # Simple change in white frequency
    delta_pW = pW_t6 - pW_t0,

    # Logit-scale change
    delta_logit_pW = qlogis(pW_t6) - qlogis(pW_t0),

    change_direction = case_when(
      delta_pW > 0 ~ "increase",
      delta_pW < 0 ~ "decrease",
      TRUE ~ "no_change"
    )
  ) %>%
  arrange(group_number, white_strain, predationYN, replicate)

# ============================================================
# QC output
# ============================================================
cat("\nRows in joined dataset:", nrow(fitness_white_change), "\n")
cat("Rows with both timepoints:", sum(fitness_white_change$has_both_timepoints, na.rm = TRUE), "\n")
cat("Rows passing min_colony_count filter:", sum(fitness_white_change$passes_min_colonies, na.rm = TRUE), "\n")

cat("\nRows missing t0 match:\n")
print(
  fitness_white_change %>%
    filter(!has_both_timepoints) %>%
    select(group_number, white_strain, replicate, predationYN,
           num_blue_colonies_t0, num_white_colonies_t0,
           num_blue_colonies_t6, num_white_colonies_t6),
  n = 50
)

cat("\nRows failing min_colony_count filter:\n")
print(
  fitness_white_change %>%
    filter(has_both_timepoints, !passes_min_colonies) %>%
    select(group_number, white_strain, replicate, predationYN,
           total_count_t0_raw, total_count_t6_raw),
  n = 50
)
```

```{r}
# ============================================================
# Save full paired data table
# ============================================================
readr::write_csv(
  fitness_white_change,
  file.path(outdir, "white_frequency_change_t0_to_t6_by_group_predation_replicate.csv")
)
```

```{r}
# ============================================================
# Analysis dataset = only rows with both timepoints and enough colonies
# ============================================================
analysis_df =
  fitness_white_change %>%
  filter(passes_min_colonies)

# ============================================================
# Summary table by white strain and predation treatment
# ============================================================
fitness_white_summary =
  analysis_df %>%
  group_by(white_strain, predationYN) %>%
  summarise(
    n = n(),
    mean_pW_t0 = mean(pW_t0, na.rm = TRUE),
    mean_pW_t6 = mean(pW_t6, na.rm = TRUE),
    mean_delta_pW = mean(delta_pW, na.rm = TRUE),
    sd_delta_pW = sd(delta_pW, na.rm = TRUE),
    se_delta_pW = sd_delta_pW / sqrt(n),
    mean_delta_logit_pW = mean(delta_logit_pW, na.rm = TRUE),
    sd_delta_logit_pW = sd(delta_logit_pW, na.rm = TRUE),
    se_delta_logit_pW = sd_delta_logit_pW / sqrt(n),
    ci95_lo_delta_logit = mean_delta_logit_pW - 1.96 * se_delta_logit_pW,
    ci95_hi_delta_logit = mean_delta_logit_pW + 1.96 * se_delta_logit_pW,
    .groups = "drop"
  ) %>%
  mutate(
    # one-sided error bars away from zero
    err_ymin = if_else(mean_delta_logit_pW >= 0,
                       mean_delta_logit_pW,
                       mean_delta_logit_pW - 1.96 * se_delta_logit_pW),
    err_ymax = if_else(mean_delta_logit_pW >= 0,
                       mean_delta_logit_pW + 1.96 * se_delta_logit_pW,
                       mean_delta_logit_pW)
  ) %>%
  arrange(white_strain, predationYN)

print(fitness_white_summary, n = 100)

readr::write_csv(
  fitness_white_summary,
  file.path(outdir, "white_strain_fitness_summary_t0_to_t6_by_predation.csv")
)
```


# Plot 1: paired plot for each white_strain
```{r} 
plot_df =
  analysis_df %>%
  mutate(pair_id = row_number()) %>%
  select(
    pair_id,
    group_number,
    white_strain,
    replicate,
    predationYN,
    pW_t0,
    pW_t6,
    change_direction
  ) %>%
  pivot_longer(
    cols = c(pW_t0, pW_t6),
    names_to = "timepoint",
    values_to = "pW"
  ) %>%
  mutate(
    timepoint = recode(
      timepoint,
      pW_t0 = "start",
      pW_t6 = "end"
    ),
    timepoint = factor(timepoint, levels = c("start", "end")),
    predationYN = factor(predationYN, levels = c("no", "yes"))
  )

# reorder the levels of white_strain
plot_df$white_strain <- factor(plot_df$white_strain, levels = c("ancW", paste0("ev", 1:12)))

p1 =
  ggplot(plot_df, aes(x = timepoint, y = pW, group = pair_id, color = predationYN)) +
  geom_line(alpha = 0.5) +
  geom_point(size = 2, alpha=0.6) +
  facet_wrap(~ white_strain) +
  scale_color_manual(values = predation_colors, drop = FALSE) +
  theme_classic() +
  labs(
    x = NULL,
    y = "White strain frequency",
    color = "Predation",
    title = "Change in white strain frequency from start to end"
  )

print(p1)


ggsave(
  filename = file.path(outdir, "paired_white_frequency_by_white_strain.pdf"),
  plot = p1,
  width = 10,
  height = 7
)
```

```{r}
# without predation
p_noPred = ggplot(data=fitness_white_change %>% filter(predationYN=="no"), aes(x=white_strain, y=delta_pW)) +
  geom_boxplot(aes(x=white_strain, fill=predationYN), alpha=0.5, outlier.shape=NA) +
  geom_jitter(width=0.1, alpha=0.6) + 
  geom_hline(yintercept=0, lty=2) +
  #scale_color_manual(values=predation_colors) +
  scale_fill_manual(values=predation_colors) +
  ylab("Change in Frequency") +
  theme_cowplot() +ylim(-1,1) + xlab("White Strain") + ggtitle("No Predation")


# with predation
p_wPred = ggplot(data=fitness_white_change %>% filter(predationYN=="yes"), aes(x=white_strain, y=delta_pW)) +
  geom_boxplot(aes(x=white_strain, fill=predationYN), alpha=0.5, outlier.shape=NA) +
  geom_jitter(width=0.1, alpha=0.6) + 
  geom_hline(yintercept=0, lty=2) +
  #scale_color_manual(values=predation_colors) +
  scale_fill_manual(values=predation_colors) +
  ylab("Change in Frequency") +
  theme_cowplot() +ylim(-1,1) + xlab("White Strain") + ggtitle("With Amoebal Predation")

p_all=plot_grid(p_noPred, p_wPred, ncol=1)

save_plot("../output/with_vs_without_predation.pdf", p_all, base_height = 8, base_width=8)


```


# Plot 2: paired plot separated by predation treatment
```{r}

p2 =
  ggplot(plot_df, aes(x = timepoint, y = pW, group = pair_id, color = predationYN)) +
  geom_line(alpha = 0.5) +
  geom_point(size = 2, alpha=0.7) +
  facet_grid(predationYN ~ white_strain) +
  scale_color_manual(values = predation_colors, drop = FALSE) +
  theme_classic() +
  labs(
    x = NULL,
    y = "White strain frequency",
    color = "Predation",
    title = "Change in white strain frequency from start to end by predation treatment"
  )

print(p2)

ggsave(
  filename = file.path(outdir, "paired_white_frequency_by_white_strain_and_predation.pdf"),
  plot = p2,
  width = 12,
  height = 8
)
```

# Plot 3: summary plot of mean delta_logit_pW
# one-sided error bars away from zero

```{r} 
fitness_white_summary =
  fitness_white_summary %>%
  mutate(predationYN = factor(predationYN, levels = c("no", "yes"))) 
p3 =
  ggplot(fitness_white_summary,
         aes(x = white_strain, y = mean_delta_pW, fill = predationYN)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_errorbar(aes(ymin = mean_delta_pW+se_delta_pW, ymax = mean_delta_pW-se_delta_pW),
    position = position_dodge(width = 0.8),width = 0.2) +
  scale_fill_manual(values = predation_colors, drop = FALSE) +
  theme_classic() +
  labs(
    x = "White strain",
    y = "Mean change in logit(pW)",
    fill = "Predation",
    title = "Mean competitive change in white strain frequency"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) + geom_hline(yintercept=0, lty=2)

print(p3)

ggsave(
  filename = file.path(outdir, "summary_delta_logit_pW_by_white_strain_and_predation.pdf"),
  plot = p3,
  width = 10,
  height = 6
)
```

# QC table showing how many joined rows per t0 pairing group
```{r}

pairing_qc =
  fitness_white_change %>%
  count(group_number, white_strain, replicate, name = "n_joined_rows") %>%
  arrange(desc(n_joined_rows), group_number, white_strain, replicate)

print(pairing_qc, n = 100)

readr::write_csv(
  pairing_qc,
  file.path(outdir, "pairing_qc_counts.csv")
)


```

